Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
litter section

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
This guidance has improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC guidelines, @cook-patton_mapping_2020 found that the latest default rates may underestimate rates of C accumulation in regrowth forests by 32% on average and fail to capture eight-fold variation within ecozones. 
In addition, @cuni-sanchez_high_2021 found that aboveground C stocks in mature African tropical montane forests were two-thirds higher than the IPCC default values for these forests.
This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_202] implies that it is useful for practitioners to have access to locally-specific information, when available.

@@

Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].
IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs]. 
Beyond the minimum DBH sampled, forest censuses and biomass estimates also differ in their inclusion of life forms other than dicot trees -- including lianas, ferns, palms, and bamboo -- which in some places can reach large sizes and/or constitute a large fraction of forest C.
*[explain IPCC guidance]*
Further, it is important to note that this excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

@@

Belowground biomass is defined as "all biomass of live roots" [@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain.
Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process. 
Field methods for estimating root biomass are highly variable.
IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].

@@

Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_2006_2006]. 
This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
While dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance; @carmona_coarse_2002], aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
In turn, they are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

Litter is defined as including " all  non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil. This includes litter (OL),  fumic (OF),  and  humic (OH) layers.  Live  fine  roots  (of  less  than  the  suggested  diameter  limit for  belowground  biomass) are included in litter where they cannot be distinguished from it empirically." (2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

@@

Soil organic matter is defined as "Includes organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and
applied consistently through the time series. Live fine roots (of less than the suggested diameter limit for belowground biomass) are included with soil organic matter where they cannot be distinguished from it empirically."(2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)

@@

library(knitr)
library(kableExtra)
knitr::kable(C_variables_table, format = "latex",
             longtable = TRUE,
             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1. ") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%
  row_spec(0, bold = TRUE)  %>%
  row_spec(1, bold = TRUE)  %>% #biomass
  row_spec(5, bold = TRUE)  %>% #agb
  row_spec(13, bold = TRUE) %>% #root
  row_spec(21, bold = TRUE)  %>% #dead wood
  row_spec(28, bold = TRUE)  %>% #litter
  row_spec(33, bold = TRUE)  %>% #som
  row_spec(37, bold = TRUE) #total
```

>>>>

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
This guidance has improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC guidelines, @cook-patton_mapping_2020 found that the latest default rates may underestimate rates of C accumulation in regrowth forests by 32% on average and fail to capture eight-fold variation within ecozones. 
In addition, @cuni-sanchez_high_2021 found that aboveground C stocks in mature African tropical montane forests were two-thirds higher than the IPCC default values for these forests.
This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.

@@

Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs]. 
Beyond the minimum DBH sampled, forest censuses and biomass estimates also differ in their inclusion of life forms other than dicot trees -- including lianas, ferns, palms, and bamboo -- which in some places can reach large sizes and/or constitute a large fraction of forest C.
*[explain IPCC guidance]*
Further, it is important to note that this excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

@@

Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain.
Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process. 
Field methods for estimating root biomass are highly variable.
IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].

@@

Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_good_2003;@ipcc_2006_2006]. 
This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
While dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance; @carmona_coarse_2002], aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
In turn, they are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
As noted above, live fine roots may be included in litter when difficult to separate empirically.
The definition includes the entire O horizion, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

@@

Soil organic matter is defined as "Includes organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and
applied consistently through the time series. Live fine roots (of less than the suggested diameter limit for belowground biomass) are included with soil organic matter where they cannot be distinguished from it empirically." [@ipcc_good_2003;@ipcc_2006_2006].

@@

library(knitr)
library(kableExtra)
knitr::kable(C_variables_table, format = "latex",
             longtable = TRUE,
             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1. ") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%
  row_spec(0, bold = TRUE)  %>%
  row_spec(1, bold = TRUE)  %>% #biomass
  row_spec(5, bold = TRUE)  %>% #agb
  row_spec(13, bold = TRUE) %>% #root
  row_spec(21, bold = TRUE)  %>% #dead wood
  row_spec(28, bold = TRUE)  %>% #litter
  row_spec(35, bold = TRUE)  %>% #som
  row_spec(39, bold = TRUE) #total
```
